---
title: "Boreal Caribou Simulations Analysis"
author: "Dalgarno, S., Thorley, J.L., Boulanger, J. & Pearson, A."
bibliography: bibliography.bib
---
  
```{r, echo = FALSE, warning = FALSE, message = FALSE, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE, comment = NA, results = "asis", cache = FALSE)

figure_tag <- function(file, precap){
  path <- file.path("figures/simulations/plots", file)
  rds <- readRDS(paste0(path, ".rds"))
  rda <- readRDS(paste0(path, ".rda"))
  png <- paste0(path, ".png")
  width <- rda$width/6 * 100
  caption <- rda$caption
  img <- paste0("<img alt = \"", png, "\" src = \"", png, "\" title = \"", 
      png, "\" width = \"", width, "%\">")
  caption <- paste0("<figcaption>", precap, " ", caption, "</figcaption>")
  paste0(c(character(0), "<figure>", img, caption, "</figure>"), collapse = "\n")
}

table_tag <- function(file, precap){
  path <- file.path("figures/simulations/tables", file)
  rds <- readRDS(paste0(path, ".rds"))
  rda <- readRDS(paste0(path, ".rda"))
  caption <- rda$caption
  caption <- paste0(precap, " ", caption)
  table <- knitr::kable(rds, format = "markdown", row.names = FALSE)
  txt <- c(caption, "", table)
  txt <- c(txt, "")
  paste0(txt, collapse = "\n")
}
```

## Background

The objectives of this analysis are to:  

- Simulate Boreal Caribou recruitment and survival data from key demographic and sampling parameters.  
- Assess the ability of various statistical methods to recover known parameter values from simulated data.

One hundred recruitment and survival data sets were simulated from a single set of demographic (table 1) and sampling (table 2) parameters for each of 5 coverage scenarios (table 3). 
Coverage scenarios included variation in the number of collared adult females (for survival data) and the proportion of groups observed in hypothetical composition surveys (for recruitment data).
Realistic parameter values for survival/fecundity rates and population simulation were derived from observed values in anonymized example datasets.
The simulated population is stable (i.e., lambda close to 1; table 4). 
See [this bbousims vignette](https://poissonconsulting.github.io/bbousims/articles/bbousims.html) for more information on how survival and recruitment data are simulated, given the parameter values in table 1 and table 2. 

Statistical methods compared include a Bayesian model with year as fixed effect, a Bayesian model with year as random effect, and a frequentist retro model. 
See [this bboutools vignette](https://poissonconsulting.github.io/bboutools/articles/methods.html) for more details on the statistical methods.
All Bayesian models had uninformative priors. 

Performance metrics used include bias, precision and coverage.
Bias is a measure of the degree to which the estimate over- or under-estimates the true value. 
An unbiased estimator is one for which the mean bias across all estimates is zero. 
Precision is simply the difference between the upper and lower limits of the 95% CI.
Note that for Bayesian models CI refers to the credible interval, whereas CI represents the confidence interval in frequentist methods. 
Coverage is the proportion of estimates for which the 95% CI contain the true value.
All performance metrics are visualized as boxplots, which display the median and 95% interquartile range. 
Each boxplot is a summary of 2000 points (i.e., 100 simulations by 20 years).

Frequentist retro methods fail to estimate standard error (and 95% CI) for annual survival when there are 0 observed mortalities. 
In these cases, the annual population growth (lambda) estimates are also missing 95% CI. 
It is important to note that the precision plots exclude these values and the coverage plots assume that the known value is not within the 95% CI. 
The number (and proportion) of these cases is summarized for each coverage scenario in table 5 and 6. 
  
## Results

### Population
```{r}
cat(table_tag("params_rates", precap = "Table 1."))
cat(table_tag("params_pop", precap = "Table 2."))
cat(table_tag("scenarios", precap = "Table 3."))
cat(table_tag("demographic_summary", precap = "Table 4."))
cat(figure_tag("population", precap = "Figure 1."))
```

### Survival
```{r}
cat(figure_tag("survival/bias", precap = "Figure 2."))
cat(figure_tag("survival/precision", precap = "Figure 3."))
cat(figure_tag("survival/coverage", precap = "Figure 4."))
cat(table_tag("survival/n_se0", precap = "Table 5."))
```

### Recruitment
```{r}
cat(figure_tag("recruitment/bias", precap = "Figure 5."))
cat(figure_tag("recruitment/precision", precap = "Figure 6."))
cat(figure_tag("recruitment/coverage", precap = "Figure 7."))
```

### Growth
```{r}
cat(figure_tag("growth/bias", precap = "Figure 8."))
cat(figure_tag("growth/precision", precap = "Figure 9."))
cat(figure_tag("growth/coverage", precap = "Figure 10."))
cat(table_tag("growth/n_noCI", precap = "Table 6."))
```

## Discussion
For the survival model, higher sample size resulted in reduced bias, higher precision and higher coverage for all statistical methods.
The Bayesian random effects model performed best, with improvement over other methods particularly at lower sample sizes. 

Results for the recruitment model were less clear, with methods performing similarly across sample sizes.
Surprisingly, coverage was reduced with higher proportion of groups observed in the Bayesian recruitment models. 
We suspect that this is caused by higher precision leading to a higher tendency for the 95% CI to not include the true value. 
Considering population growth estimates generally, the Bayesian random effects model performed best.

The growth predictions are derived from the recruitment and survival estimates and thus, it is unsurprising that the Bayesian random effects model performed best. 

A consequence of modeling year as a random effect is that individual estimates are pulled in toward the grand mean, especially when estimates have low precision or are more extreme [@kery_bayesian_2011].
This simulation analysis demonstrates that this 'shrinkage' is desirable as the random effects model tends to be more skeptical of extreme values. 
Extreme values are likely to result from the sampling process rather than the ecological processes, especially at low sample sizes.

